---
title: 'Lab 2: Intro to Shiny'
author: "Julianna Renzi"
date: "1/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(janitor) # simple tools for examining/cleaning dirty data
library(palmerpenguins)
library(here)

# for PCA
library(ggfortify)

# for ggplot customation
library(readxl) # read in Excels
library(gghighlight) # color things by group
library(lubridate) # dates
library(plotly) # interactive graphs

# to make compound figures
library(patchwork)
```

## Part 1: PCA with penguins

```{r}
# use view(penguins) to see whole dataset
  # see lots have NAs
  # also need to scale
penguin_pca <- penguins %>% 
  select(body_mass_g, ends_with("_mm")) %>%  # need contiguous variables only; want all mm functions
  drop_na() %>% # any row with any NA (listwise deletion if there are any NAs)--add variables in drop_na() if you want to drop rows with NAs in particular columns 
  scale() %>% # scales variables
  prcomp() # run PCA--now it's not a df it's a list with information on the PCA

penguin_pca$rotation
  # loadings for each variable on each PC
  # axis rotation that is a linear combination of the original variables

# create a biplot (could do it manually in ggplot, but this is easier)
# autoplot guesses the plot you need (needs ggfortify)--"appropriate to a particular data type" i.e. assumes you want a PCA biplot
autoplot(penguin_pca) # this is just the locations in multivariate space on PC1 and PC2

# for aesthetics want a dataset with the same observations that were used to create the PCA for coloring, etc.
penguin_complete <- penguins %>% 
  drop_na(body_mass_g, ends_with("_mm")) # don't need to worry about columns, just rows (also don't want to drop all na because this still has other columns with different NAs)
# want all the columns because might want to color based on a factor (not a continuous thing that was used)

```

```{r}
autoplot(penguin_pca,
         data = penguin_complete,
         colour = "species",
         loadings = TRUE,
         loadings.label = TRUE) +
  theme_minimal()
# the data MUST match (observations/rows are exactly the same as PCA even if columns are different)
# see Gentoos (bigger) are much bigger than Adelie and Chinstrap--explains 90% of the data!!
# look at example code for other nice formatting ideas
# defactoextra? factoextra? one is good for even better PCA visualizations

```

## Part 2: ggplot2 customization and reading in different file types

```{r}
# read in Excel file to wrangle and graph
fish_noaa <- read_excel(here("data", "foss_landings.xlsx"))  %>% 
  clean_names() %>% # from janitor: convert to lower snake case?
  mutate(across(where(is.character), tolower)) %>% # convert character columns to lower case (easier to type)
# across = across columns 1:5, or name:year, or all characters--could also do ends_with, starts_with, etc.
  mutate(nmfs_name = str_sub(nmfs_name, end = -4)) %>% # get rid of asteriks and spaces (last 4 characters)
# str_sub
  filter(confidentiality == "public")
  # only keep observations that are public

```

Make a customized graph:

```{r}
fish_plot <- ggplot(data = fish_noaa,
                    aes(x = year, y = pounds)) + 
  geom_line(aes(color = nmfs_name), show.legend = FALSE) + 
  theme_minimal()

fish_plot # so many species that only see the legend!
```

Make it interactive so can look at all of them
```{r}
ggplotly(fish_plot) # now can hover over different series to get info
# can drag cursor over an area to zoom in
# don't need to do it all the time, but is cool (interactive in knitted form also)

```


```{r}
# this is a sphagetti plot becuase there are SO many lines on it, but sometimes want to highlight a few using gghighlight

ggplot(data = fish_noaa,
                    aes(x = year, y = pounds, group = nmfs_name)) + 
  geom_line() + 
  theme_minimal() +
  gghighlight(nmfs_name == "tunas") # all other series in gray, black is one of interest

# can also highlight based on values
ggplot(data = fish_noaa,
                    aes(x = year, y = pounds, group = nmfs_name)) + 
  geom_line(aes(color = nmfs_name)) + 
  theme_minimal() +
  gghighlight(max(pounds) > 1e8) # but both in black auto so need to put color in geom_line() and gghighlight takes it away !! what an MVP

```

## Part 3: Read in data from a URL, use 'lubridate()' to work with dates/times, use 'mutate()' a bunch, and make a graph in logical order

Wrangling using Mutate again and changing factor level to show up in a reasonable order. Data from the Monroe water treatment plant daily electricity use (copied from online) 

```{r}
# most updated version (but can change your outputs)

monroe_wt <- read_csv("https://data.bloomington.in.gov/dataset/2c81cfe3-62c2-46ed-8fcf-83c1880301d1/resource/13c8f7aa-af51-4008-80a9-56415c7c931e/download/mwtp_dailyelectricity_bclear.csv") %>% # still a CSV, just from a URL
  clean_names() # change column titles to lower snake case with underscores between capitalized things

```
```{r}
# convert date to a date class, then pull month, then make a new column that has the month abbreviation, then convert month to factor where order is based on month order associated with it (logical order)
monroe_ts <- monroe_wt %>% 
  mutate(date = mdy(date)) %>% # lubridate form
  mutate(record_month = month(date)) %>% 
  mutate(month_name = month.abb[record_month]) %>% # get abbreviations
  mutate(month_name = fct_reorder(month_name, record_month)) # set order based on factors in another variable (i.e. set month_name in the order of record_month)

# see this worked:
levels(monroe_ts$month_name)
  
# note:
month.abb[4] # this is how that works
month.name[4]

# plot
ggplot(data = monroe_ts, aes(month_name, y = total_k_wh)) +
  geom_jitter()

```

## Part 4: Compound figures with `patchwork`
Compound figures (i.e. multiple graphs in one figure)

```{r}
graph_a <- ggplot(data = penguins, aes(x = body_mass_g, y = flipper_length_mm)) + 
  geom_point()

graph_b <- ggplot(data = penguins, aes(x = species, y = flipper_length_mm)) + 
  geom_jitter(aes(color = species), show.legend = FALSE)

graph_a
graph_b

```
```{r}
# Use | to put graphs side by side
# use / to put graphs one over the other
graph_c <- (graph_a | graph_b) / fish_plot & theme_dark()

# can graphics as one with the & sign
# can also customize titles, layout, blank spaces, labels, etc.
# can also combine tables, etc.
graph_c

# export most recent graph
ggsave(here("fig", "graph_c_ah.png"), width = 5, height = 6) # height/width in inches
```
 

